# Vectors {#vectors}

```{r, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', collapse = TRUE, R.options = list(width = 80))
```

## Aims

* Learning how to work with R code files
* Getting to know the simplest data structure in R, the vector
* Learning about subsetting, one of the fundamental operations with data

## Editing code

### Using code files

* Computer code is stored as **plain text**
* When writing computer code, we must use a **plain text editor**, such as **Notepad++** or **RStudio**
* A **word processor**, such as **Microsoft Word**, is not a good choice for writing code, because - 
    * Documents created with a Word Processor contain elements **other than** plain text (such as formatting), which are not processed, leading to confusion
    * Word processors can automatically **correct** "mistakes" thereby introducing unintended changes in our code, such as capitalizing: `max(1)` → `Max(1)`

* Any plain text file can be used to store **R code** 
* Conventionally, R code files have the `*.R` file **extension**
* **Complete** code files can be executed with `source`
* Selected **parts** of code can be executed by marking the section and pressing **Ctrl+Enter**
* Executing **one expression** can be done by standing inside it with the marker and pressing **Ctrl+Enter**

```{r executing-r-code, fig.cap="Methods of executing R code", echo=FALSE, out.width="100%", auto.pdf=TRUE}
knitr::include_graphics("images/lesson_02_source.svg")
```

For example, we can run the file `volcano.R` which draws a 3D image of a volcano (Figure \@ref(fig:image-volcano-3d)). 

```{r image-volcano-3d, fig.cap="3D image of the `volcano` dataset", fig.align="center"}
source("data/volcano.R")
```

### RStudio keyboard stortcuts

RStudio has numerous keyboard shortcuts for making it easier to edit and execute code files. Some useful RStudio keyboard shortcuts are given in Table \@ref(tab:rstudio-shortcuts). 

| Shortcut | Action | 
|---|---|
| **Alt+Shift+K** | List of all shortcuts |
| **Ctrl+1** | Moving cursor to the code editor | 
| **Ctrl+2** | Moving cursor to the console |
| **Ctrl+Enter** | Sending the current selection or line |
| **Ctrl+Shift+P** | Re-sending last selection
| **Ctrl+Alt+B** | Sending from top to current line |
| **Ctrl+Shift+C** | Turn comment on or off |
| **Tab** | Auto-complete |
| **Ctrl+D** | Delete line |
| **Ctrl+Shift+D** | Duplicate line |
| **Ctrl+F** | Find and replace menu |
| **Ctrl+S** | Save | 

Table: (\#tab:rstudio-shortcuts) RStudio keyboard shortcuts

## Assignment

* So far we have been using R by **typing** expressions into the command line and observing the result on screen
* That way R functions as a **"calculator"** - the results are not kept in computer memory

```{r, echo=FALSE, out.width="100%", auto.pdf=TRUE}
knitr::include_graphics("images/lesson_02_print.svg")
```

* Storing objects in the temporary computer memory (RAM) is called **assignment**
* Assignment is done using the **assignment operator** 

* In an **assignment expression** we are storing an object, under a certain name, in the RAM
* This is an essential operation in programming, because it make **automation** possible - reaching the goal step by step, while storing intermediate products
* An assignment expression consists of - 
    * The **expression** whose result we want to store
    * The assignment **operator**, `=` or `<-`
    * The **name** which will be assigned to the object
* For example - 

```{r}
rateEstimate = (6617747987 - 6617746521) / 10
```

```{r, echo=FALSE, out.width="100%", auto.pdf=TRUE}
knitr::include_graphics("images/lesson_02_assignment1.svg")
```

* When we type an **object name** in the console, R **accesses** an object stored under that name in the RAM, and **calls** the `print` function on the object - 

```{r}
rateEstimate
```

```{r}
print(rateEstimate)
```

```{r, echo=FALSE, out.width="100%", auto.pdf=TRUE}
knitr::include_graphics("images/lesson_02_assignment2.svg")
```

* What happens when we assign a new value to an **existing object**? The old object gets deleted, and its name is now pointing on the new value

```{r}
x = 55
x
```

```{r}
x = "Hello"
x
```

* Note the difference between the `==` and `=` operators!
* `=` is an **assignment** operator - 

```{r}
one = 1
two = 2
one = two
one
two
```

* `==` is a logical operator for **comparison** - 

```{r}
one = 1
two = 2
one == two
```

* Which user defined objects are currently **in memory**? The `ls` function returns a character vector with their names - 

```{r, eval=FALSE}
ls()
```

* Question: why do we write `ls()` and not `ls`?

## Vectors

### What is a vector?

* A vector is an **ordered** collection of values of the **same type**, such as - 
    * **Numbers** - `interger` or `numeric`
    * **Text** - `character`
    * **Logical** - `logical`
* Recall that these are the same three types of constant values we saw in **Lesson 01**; in fact a constant value is a vector of length 1
* We can create an **empty vector** using the `vector` function, specifying - 
    * `mode` - Type
    * `length` - Number of elements

```{r}
v = vector(mode = "numeric", length = 10)
v
```

### The `c` function

* Vectors can also be created with the `c` function, which **combines** the given vectors in the given order - 

```{r}
x = c(1, 2, 3)
x
```

```{r}
c(x, 5)
```

* Another example, with `character` values - 

```{r}
y = c("cat", "dog", "mouse", "apple")
y
```

### Vector subsetting

* We can access individual vector **elements** using the `[` operator and an index; in other words, to get a subset with an individual vector element -

```{r}
y[2]
```

```{r}
y[3]
```

* Note: the index starts at `1`!

* Another example - 

```{r}
counts = c(2, 0, 3, 1, 3, 2, 9, 0, 2, 1, 11, 2)
counts[4]
```

* Components of an expression for accessing a vector element - 

```{r, echo=FALSE, out.width="50%", auto.pdf=TRUE}
knitr::include_graphics("images/lesson_02_subset1.svg")
```

* We can also **assign** new values into a vector subset - 

```{r}
x = c(1, 2, 3)
x
```

```{r}
x[2] = 300
x
```

* Note: in this example we made an assignment into a subset with a single element; the same way, we can make an assignment into a subset of any length (see below)

### Calling functions on a vector

* There are various functions for calculating vector **properties** - 

```{r}
x = c(1, 6, 3, -8, 2)
```

```{r}
length(x)  # Number of elements
min(x)     # Minimum
max(x)     # Maximum
range(x)   # Minimum, maximum
mean(x)    # Average
sum(x)     # Sum
```

* Other functions operate on **each** vector element, returning a vector of results having the same length as the input - 

```{r, warning=FALSE}
sqrt(x)
```

```{r}
abs(x)
```

* Question: why does the result of the first expression contain `NaN`?

### The recycling rule

* Binary operations applied on two vectors are done **element-by-element**, and a vector of the results is returned - 

```{r}
c(1, 2, 3) * c(10, 20, 30)
```

* What happens when the input vector lengths do not match? The shorter vector gets **"recycled"**

* For example, when one of the vectors is of length 3 and the other vector is of length 6, then the sorter (of length 3) is repeated 2 times until it matches the longer vector - 

```{r}
c(1, 2, 3)          + c(1, 2, 3, 4, 5, 6)
```

```{r}
c(1, 2, 3, 1, 2, 3) + c(1, 2, 3, 4, 5, 6)
```

```{r, echo=FALSE, out.width="50%", auto.pdf=TRUE}
knitr::include_graphics("images/lesson_02_recycle.svg")
```

* When one of the vectors is of length 1 and the other is of length 4, the shorter vector (of length 1) is replicated 4 times - 

```{r}
2             * c(1, 2, 3, 4)
```

```{r}
c(2, 2, 2, 2) * c(1, 2, 3, 4)
```

* When one of the vectors is of length 2 and the other is of length 4, the shorter vector (of length 2) is replicated 2 times - 

```{r}
c(10, 100)          + c(1, 2, 3, 4)

```

```{r}
c(10, 100, 10, 100) + c(1, 2, 3, 4)

```

* When longer vector length is not a multiple of the shorter one recycling is **"incomplete"** and we get a warning message -

```{r}
c(1, 2)    * c(1, 2, 3)
```

```{r}
c(1, 2, 1) * c(1, 2, 3)
```

* The recycling rule applies to many operators and functions, such as all arithmetic operators -

```{r}
x = c(1, 2, 3)
y = c(4, 5, 6)
```

```{r}
x + y
```

```{r}
x - y
```

```{r}
x * y
```

```{r}
x / y
```

### Consecutive and repetitive vectors

* Other than the `vector` and `c` functions, there are three commonly used methods for creating **consecutive** or **repetitive** vectors - 
    * The `:` operator
    * The `seq` function
    * The `rep` function

#### Consecutive vectors

* The `:` operator is used to create a vector of consecutive vectors in steps of `1` or `-1` - 

```{r}
1:10   # Steps of 1
```

```{r}
55:43  # Steps of -1
```

* The `seq` function creates a vector of consecutive values in any step size - 
    * `from` - Where to start
    * `to` - When to end
    * `by` - Step size
* For example - 

```{r}
seq(from = 100, to = 150, by = 10)
```

```{r}
seq(from = 100, to = 80, by = -5)
```

#### Repetitive vectors

* The `rep` function **replicates** its argument to create a repetitive vector - 
    * `x` - What to replicate
    * `times` - How many time to repeat `x`
    * `each` - How many times to repeat each element of `x`

```{r}
rep(x = 22, times = 10)
```

```{r}
rep(x = 22, each = 10)
```

* The `x` argument can be a vector of **length >1**
* Note the difference between the `times` and the `each` **parameters** - 

```{r}
rep(x = c(18, 0, 9), times = 3)
```

```{r}
rep(x = c(18, 0, 9), each = 3)
```

### Function calls

* Using the `seq` function we will demonstrate three properties of function calls
* **First**, we can omit parameter names as long as the arguments are passed in the default order - 

```{r}
seq(from = 5, to = 10, by = 1)
```

```{r}
seq(5, 10, 1)
```

* **Second**, we can use any argument order as long as parameter names are specified - 

```{r}
seq(to = 10, by = 1, from = 5)
```

```{r}
seq(by = 1, from = 5, to = 10)
```

```{r}
seq(from = 5, by = 1, to = 10)
```

* **Third**, we can omit parameters that have a default value in the function definition - 

```{r}
seq(5, 10, 1)
```

```{r}
seq(5, 10)
```

* To find out what are the parameters of a particular function, their order or their default values, we can look into the documentation - 

```{r}
# ?seq
```

### Vector subsetting

* So far we created vector subsets using a `numeric` index which consists of a **single** value, such as - 

```{r}
x = c(43, 85, 10)
x[3]
```

* We can also use a vector of **length >1** as an index - 

```{r}
x[1:2]
```

* Note that the vector does not need to be consecutive, and can include repetitions - 

```{r}
x[c(1, 1, 3, 2)]
```

* Another example - 

```{r}
counts = c(2, 0, 3, 1, 3)
counts[1:3]
```

```{r, echo=FALSE, out.width="50%", auto.pdf=TRUE}
knitr::include_graphics("images/lesson_02_subset2.svg")
```

* Another example - 

```{r}
counts = c(2, 0, 3, 1, 3, 2, 9, 0, 2, 1, 11, 2)
counts[c(1:3, 7:9)]
```

```{r, echo=FALSE, out.width="75%"}
knitr::include_graphics("images/lesson_02_subset3.svg")
```

* For the next examples, let's create a vector of all **even** numbers between 1 and 100 -

```{r}
x = seq(2, 100, 2)
x
```

* Question: what is the meaning of the numbers in square brackets when printing the vector?

* How many **elements** does `x` have?

```{r}
length(x)
```

* What is the value of the **last** element in `x`?

```{r}
x[50]
```

```{r}
x[length(x)]
```

* Question: which of the last two expressions is preferable and why?

* How can we get the **entire** vector using subsetting with a `numeric` index?

```{r}
x[1:length(x)]
```

* How can we get the entire vector **except for** the last element?

```{r}
x[1:(length(x)-1)]
```

* Question: how can we get a reversed vector using a `numeric` index?

* When requesting an index **beyond** vector length we get `NA` -

```{r}
x[55]
```

```{r}
x[1:80]
```

* Reminder: `NA` is a special value meaning *Not Available*

### The recycling rule

* Earlier, we saw the recycling rule with **arithmetic operators**
* The rule also applies to **assignment** - 

```{r}
counts = c(2, 0, 3, 1, 3, 2, 9, 0, 2, 1, 11, 2)
counts[c(1:3, 7:9)]
```

```{r}
counts[c(1:3, 7:9)] = NA
counts
```

```{r}
counts[c(1:3, 7:9)] = c(NA, 99)
counts
```

### Logical vectors

* The third common type of vectors are `logical` vectors
* A **logical vector** is composed of logical values: `TRUE` and `FALSE`
* For example - 

```{r}
c(TRUE, FALSE, FALSE)
```

```{r}
rep(TRUE, 7)
```

* Usually, we will not be creating `logical` vectors **"manually"**, but through applying a **logical operator** on another vector -

```{r}
x = 1:5
x
```

```{r}
x >= 3
```

* Note how the recycling rule applies to logical operators as well

* When arithmetic operations are applied to a `logical` vector, the `logical` vector is **converted** to a numeric one, where `TRUE` becomes `1` and `FALSE` becomes `0`
* For example - 

```{r}
sum(x >= 3)
```

```{r}
mean(x >= 3)
```

* Question: what is the meaning of the values `3` and `0.6` in the above example?

* We can use a `logical` vector as an **index** for subsetting - 

```{r}
counts = c(2, 0, 3, 1, 3)
```

```{r}
counts < 3
```

```{r}
counts[counts < 3]
```

* The logical vector `counts<3` specifies whether **to include** each of the elements of `counts` in the resulting subset

```{r, echo=FALSE, out.width="50%", auto.pdf=TRUE}
knitr::include_graphics("images/lesson_02_subset_logical.svg")
```

* Using a `logical` index - 

```{r}
x = 1:5
```

```{r}
x[x >= 3]
```

```{r}
x[x != 2]
```

```{r}
x[x > 4 | x < 2]
```

```{r}
x[x > 4 & x < 2]
```

* Question: what happened in the last expression?

* The next example is slightly more complex: we select the elements of `z` whose square is larger than 8 - 

```{r}
z = c(5, 2, -3, 8)
z[z^2 > 8]
```

* `z^2` gives a vector of squared `z` values (`2` is recycled)

```{r}
z^2
```

* Each of the squares is compared to 8 (`8` is recycled)

```{r}
z^2 > 8
```

* Finally, the latter `logical` vector is used for subsetting `z`

### Missing values

* The `is.na` function is used to detect **missing** (`NA`) values in a vector - 
    * Accepts a vector of **any type**
    * Returns a **logical** vector with `TRUE` in place of `NA` values and `FALSE` in place of non-`NA` values
* For example -

```{r}
x = c(28, 58, NA, 31, 39, NA, 9)
x
```

```{r}
is.na(x)
```

* Many functions that **summarize** vector properties, such as `mean`, have the `na.rm` parameter
* The `na.rm` parameter can be used to **exclude** `NA` values from the calculation
* The default is `na.rm=FALSE`, meaning that `NA` values are **not** excluded
* For example - 

```{r}
x = c(28, 58, NA, 31, 39, NA, 9)
```

```{r}
mean(x)
mean(x, na.rm = TRUE)
```

* Question 1: why do we get `NA` in the first expression? 
* Question 2: what will be the result of `length(x)`?

```{r}
x = c(28, 58, NA, 31, 39, NA, 9)
x
```

* Question: how can we replace the `NA` values in the above vector with the mean of its non-`NA` values?

### The `any` and `all` functions

* Sometimes we want to figure out whether - 
    * A `logical` vector contains **at least one** `TRUE` value
    * A `logical` vector is **entirely** composed of `TRUE` values
* We can use the `any` and `all` functions, respectively, for that

* The `any` function returns `TRUE` if **at least one** of the input vector values is `TRUE`, otherwise it returns `FALSE` - 

```{r}
x = 1:7
x
```

```{r}
x > 5
any(x > 5)
```

```{r}
x > 88
any(x > 88)
```

* The `all` function returns `TRUE` if **all** of the input vector values are `TRUE`, otherwise it returns `FALSE` - 

```{r}
x = 1:7
x
```

```{r}
x > 5
all(x > 5)
```

```{r}
x > 0
all(x > 0)
```

### The `which` function

* The `which` function **converts** a `logical` vector to a `numeric` one with the indices of `TRUE` values
* That way we can find out the **index** of values that satisfy a given criterion
* For example - 

```{r}
x = c(2, 6, 2, 3, 0, 1)
x
```

```{r}
x > 2.3
```

```{r}
which(x > 2.3)
```

### The `which.min` and `which.max` functions

* Related functions `which.min` and `which.max` return the **index** of the (first) **minimal** or **maximal** value in a vector
* For example - 

```{r}
x = c(2, 6, 2, 3, 0, 1, 6)
x
```

```{r}
which.min(x)
```

```{r}
which.max(x)
```

* Question: how can we get **all** indices of the minimal or maximal value?

## The `order` function

* The `order` function returns **ordered** vector **indices**, based on the order of vector **values**
* In other words, `order` gives the index of the smallest value, the index of the second smallest value, and so on

```{r}
x = c(2, 6, 2, 3, 0, 1, 6)
x
```

```{r}
order(x)
```

* We can also get the **reverse** order with `decreasing=TRUE` - 

```{r}
order(x, decreasing = TRUE)
```

* Question: how can we *sort* a vector using `order`?

### The `paste` and `paste0` functions

* The `paste` function is used to **"paste"** text values
* Its `sep` parameter determines the separating character(s), with default `sep=" "`

```{r}
paste("There are", "5", "books.")
paste("There are", "5", "books.", sep = "_")
```

* Non-character vectors are converted to `character` before pasting - 

```{r}
paste("There are", 80, "books.")
```

* The recycling rule applies in `paste` too - 

```{r}
paste("image", 1:5, ".tif", sep = "")
```

* The `paste0` function is a shortcut for `paste` with `sep=""` - 

```{r}
paste0("image", 1:5, ".tif")
```





