# Time series and function definitions {#time-series-and-function-definitions}

```{r, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', collapse = TRUE, R.options = list(width = 80))
```

## Aims

* Working with data which represent time (date)
* Learn how to visualize our data with graphical functions
* Learn to define custom functions

## Dates

### Date and time classes in R

* In R, there are several special classes for representing **times** and **time-series** (time+data)
    * **Times**
        * `Date` 
        * `POSIXct` 
        * `POSIXlt`
    * **Time series** (time+data) 
        * `ts` 
        * `zoo` (package `zoo`)
        * `xts` (package `xts`)

### Working with `Date` objects

The simplest data structure for representing times is `Date`, used to represent **dates** (without time of day). For example, we can get the **current** date as a `Date` object with `Sys.Date`:

```{r}
x = Sys.Date()
x
```

```{r}
class(x)
```

* We can convert `character` values in the **standard** date format `YYYY-MM-DD` to `Date`, using `as.Date` -

```{r}
as.Date("2014-10-20")
```

* When the `character` values is in a **non-standard** format, we need to specify the format definition with `format`, using the various component symbols 
* Full list of format symbols in `?strptime`

| Symbol | Meaning |
|:---:|---|
| `%d` | day (`"15"`) |
| `%m` | month, numeric (`"08"`) |
| `%b` | month, 3-letter (`"Aug"`) |
| `%B` | month, full (`"August"`) |
| `%y` | Year, 2-digit (`14`) |
| `%Y` | Year, 4-digit (`2014`) |

Table: Common `Date` format components

* For example - 

```{r}
Sys.setlocale("LC_TIME", "C")
```

```{r, error=TRUE}
as.Date("07/Aug/12")
as.Date("07/Aug/12", format = "%d/%b/%y")
```

```{r, error=TRUE}
as.Date("2012-August-07")
as.Date("2012-August-07", format = "%Y-%B-%d")
```

* The opposite conversion, with `format`, lets us **extract** specific date components out of a `Date` object - 

```{r}
d = as.Date("1955-11-30")
d
```

```{r}
format(d, "%d")
```

```{r}
format(d, "%B")
```

```{r}
as.numeric(format(d, "%Y"))
```

```{r}
format(d, "%m/%Y")
```

* `Date` objects act like **numeric** vectors with respect to certain **operations** that make sense on dates
* **Logical** operators -

```{r}
Sys.Date() > as.Date("2013-01-01")
```

* **Subtraction** -

```{r}
as.Date("2013-01-01") - as.Date("2012-01-01")
```

```{r}
as.Date("2014-01-01") - as.Date("2013-01-01")
```

* Creating **sequences** with the `seq` function - 

```{r}
seq(
  from = as.Date("2018-10-14"), 
  to = as.Date("2019-01-11"), 
  by = 7
)
```

### Time series

* Let's define two `numeric` vectors: **water level** in Lake Kinneret, in **May** and in **November**, in each year during **1991-2011** -

```{r}
may = c(
  -211.92,-208.80,-208.84,-209.12,-209.01,-209.60,
  -210.24,-210.46,-211.76,-211.92,-213.13,-213.18,
  -209.74,-208.92,-209.73,-210.68,-211.10,-212.18,
  -213.26,-212.65,-212.37
)
nov = c(
  -212.79,-209.52,-209.72,-210.94,-210.85,-211.40,
  -212.01,-212.25,-213.00,-213.71,-214.78,-214.34,
  -210.93,-210.69,-211.64,-212.03,-212.60,-214.23,
  -214.33,-213.89,-213.68
)
```

* Question 1: what is the average water level in May? in November?
* Question 2: was the water level ever below `-213` (red line) in May? in November?
* Question 3: in which year or years was the water level below `-213` in May? in November?

* Question 2: was the water level ever below `-213` (red line) in May? in November?

```{r}
any(may < -213)
```

```{r}
any(nov < -213)
```

* Note: make sure there is a space between `<` and `-`, otherwise the combination is interpreted as an assignment operator `<-`

* Question 3: in which year or years was the water level below `-213` in May? in November?

```{r}
year = 1991:2011
year
```

```{r}
year[nov < -213]
```

```{r}
year[may < -213]
```

* Note: we created a logical vector `nov < -213` and used it to subset the corresponding `year` vector

A table is more natural for representing such data:

```{r}
data.frame(year, may, nov)
```

We will learn about tables in Chapter \@ref(tables-conditionals-and-loops). 

## Graphics

### Generic functions

* Some of the functions we learned about are **generic functions**
* Generic functions are functions that can accept arguments of different classes. What the function does depends on the class, according to the **method** defined for that class
* Advantages - 
    * Easier to remember function names
    * Possible to run the same code on different types of objects
* For example, `mean`, `print` and `plot` (below) are examples of generic functions
* When the `print` function gets a **vector** it prints the values, but when it gets a **raster** `RasterLayer` object its prints a summary of its properties

### Graphical functions

* The **graphical function** `plot`, given a vector, displays its values in a two dimensional plot where - 
    * Vector **indices** are on the x-axis 
    * Vector **values** are on the y-axis
* For example - 

```{r, fig.cap="Plot of the `nov` vector", out.width="100%"}
plot(nov, type = "b")
```

* Note: `type="b"` means **both** points and lines; Other useful options include - 
    * `type="p"` for points
    * `type="l"` for lines
    * `type="o"` for overplotted lines and points

* If we pass **two** vectors to `plot`, the first appears on the x-axis and the second - on the y-axis
* For example, we can put the years of water level measurement on the x-axis as follows - 

```{r, fig.cap="\\texttt{nov} as function of \\texttt{year}", out.width="100%"}
plot(year, nov, type = "b")
```

* We can add a **horizontal line** displaying the red line using `abline` with the `h` parameter - 

```{r, eval=FALSE}
plot(year, nov, type = "b")
abline(h = -213)
```

* Note that `abline` draws in an **existing** graphical device, which is initiated with `plot`

```{r, echo=FALSE, fig.cap="Adding a horizontal line with \\texttt{abline}", out.width="100%"}
plot(year, nov, type = "b")
abline(h = -213)
```

* **Additional** "layers" can be added to an existing plot using functions such as `points` and `lines`
* We can also use **graphical parameters** to specify different style for each layer, such as `col` to determine point / line color
* In addition, we will set the y-axis **limit** with `ylim` to make sure both time series are within the displayed range
* `ylim` accepts a vector of length two: the minimum and the maximum -

```{r, fig.cap="Adding a second series with \\texttt{lines}", out.width="100%"}
plot(year, nov, ylim = range(c(nov, may)), type = "b", col = "red")
lines(year, may, type = "b", col = "blue")
abline(h = -213)
```

* Finally, we can set the axis **labels** using the `xlab` and `ylab` parameters of the `plot` function - 

```{r, fig.cap="Setting axis labels", out.width="100%"}
plot(
  year, nov,
  xlab = "Year",
  ylab = "Water level (m)",
  ylim = range(c(nov, may)),
  type = "b", col = "red"
)
lines(year, may, type = "b", col = "blue")
abline(h = -213)
```

### Consecutive differences

* The `diff` function can be used to create a vector of **differences** between consecutive elements - 

```{r}
d_nov = c(NA, diff(nov))
d_nov
```

* Note: we add `NA` to match the input vector `nov`

* Now we can find out which year had the biggest water level **increase** or **decrease** - 

```{r}
year[which.max(d_nov)]
```

```{r}
year[which.min(d_nov)]
```

* Note: `which.min` and `which.max` ignore `NA` values 

## Defining custom functions

* In **Lesson 01** we learned that a function call is an instruction to execute a certain function, as in - 

```{r, eval=FALSE}
f(arg1, arg2, ...)
```

* A **function** is an object containing code, which is loaded into the RAM and can be executed with specific parameters
* So far we met function defined in the **default** R packages (e.g. `seq`)
* Later on we will use functions from **external** packages (e.g. `left_join`)
* Now we learn how to define our own **custom** functions

```{r}
add_five = function(x) {
  x_plus_five = x + 5
  return(x_plus_five)
}
```

* Function name (`add_five`)
* Assignment operator (`=`)
* Function keyword (`function`)
* Parameter(s) (`(x)`)
* Brackets (`{`)
* Code (`x_plus_five = x + 5`)
* Returned value (`return(x_plus_five)`)
* Brackets (`}`)

* The idea is that the code inside the function gets **executed** each time the function is called - 

```{r}
# Function definition
add_five = function(x) {
  x_plus_five = x + 5
  return(x_plus_five)
}
```

```{r}
# Function call, with argument 5
add_five(5)
```

```{r}
# Function call, with argument 7
add_five(7)
```

* When we make a function call, the values we pass as function arguments are assigned to **local variables** which the function code can use
* The local variables are **not accessible** in the global environment - 

```{r, error=TRUE}
x_plus_five
```

* Every function **returns** a value
* We can **assign** the returned value to a variable to keep it in memory for later use - 

```{r}
result = add_five(3)
result
```

* We can **omit** the `return` expression -

```{r, eval=FALSE}
return(x_plus_five)
```

* In which case, the returned value is the **last expression** in the function body
* We can also omit the `{` and `}` parentheses in case the code consists of a **single expression**
* Therefore we could define the `add_five` function using **shorter** code - 

```{r}
add_five = function(x) x + 5
```

* **Default** parameter values can be given in the function definition
* In case there is a default value, we can **skip** that parameter in function calls - 

```{r, error=TRUE}
add_five = function(x) x + 5
add_five()
```

```{r}
add_five = function(x = 1) x + 5
add_five()
```

* There are **no restrictions** for the classes of objects a function can accept, as long as we did not set such restrictions ourselves
* However, we get an **error** if one of the expressions in the function code is illegal given the arguments - 

```{r}
add_five(1:3)
```

```{r, error=TRUE}
add_five("one")
```

* As another example, let's define a function named `first_last` which accepts a vector and returns the **difference** between the **last** and the **first** elements - 

```{r}
first_last = function(x) {
  x[length(x)] - x[1]
}
```

```{r}
first_last(1:3)
```

```{r}
first_last(nov)
```

* Question: write a function named `modify` that accepts three arguments - 
    * `x`
    * `index`
    * `value`
* The function assigns `value` into the element at `index` index of vector `x`
* The function returns the modified vector `x`
* For example - 

```{r, include=FALSE}
modify = function(x, index, value) {
  x[index] = value
  return(x)
}
```

```{r}
v = 1:10
v
```

```{r}
modify(x = v, index = 3, value = 99)
```















