# BGU course 2019

```{r, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', collapse = TRUE, R.options = list(width = 80))
```

## Course aims

* General knowledge in programming
* Overview of spatial data processing and analysis in R

## Course topics

**Part I - Introduction to R programming**

1. The R environment
2. Vectors
3. Time series + function definition
4. Tables + conditionals and loops

**Part II - Processing and analysis of spatial data in R**

5. Matrices and rasters
6. Raster algebra
7. Vector layers
8. Geometric operations with vector layers
9. Geometric operations with rasters
10. Working with spatio-temporal data
11. Combining rasters and vector layers
12. Spatial interpolation of point data

## Course details

* **Course number**: 128.1.0043
* **Time**: Sunday 16:10-19:00
* **Place**: Building 72, room 249
* **Instructor**: Michael Dorman (dorman@post.bgu.ac.il)
* **Grading**: 6 exercises (50%) + Exam (50%)
* **Requirements** - 
    * Basic knowledge of GIS (e.g. "Intro to GIS" course)
    * Self study
* **Getting help** - 
    * Forum on **Moodle** (http://moodle2.bgu.ac.il)
    * Meetings (Monday 15:00-16:00, schedule by e-mail)

## Lecture plan

| Lesson | Date | Slides | Exercise |
|:--:|:---:|-----------|:-----------:|
| 01 | 2019-10-27 | [**The R environment**](the-r-environment.html) | |
| 02 | 2019-11-03 | [**Vectors**](exercises/vectors.html) | [**Exercise 1**](exercises/exercise_01.pdf) | 
| 03 | 2019-11-10 | [**Time series and function definitions**](time-series-and-function-definitions.html) | | 
| 04 | 2019-11-17 | [**Tables, conditionals and loops**](tables-conditionals-and-loops.html) | [**Exercise 2**](exercises/exercise_02.pdf) | 
| 05 | 2019-11-24 | [**Matrices and rasters**](matrices-and-rasters.html) | | 
| 06 | 2019-12-01 | [**Raster algebra**](raster-algebra.html) | [**Exercise 3**](exercises/exercise_03.pdf) | 
| 07 | 2019-12-08 | [**Vector layers**](vector-layers.html) | | 
| 08 | 2019-12-15 | [**Geometric operations with vector layers**](geometric-operations-with-vector-layers.html) | [**Exercise 4**](exercises/exercise_04.pdf) | 
| 09 | 2019-12-22 | [**Geometric operations with rasters**](geometric-operations-with-rasters.html) | | 
| 10 | 2020-01-05 | [**Working with spatio-temporal data**](processing-spatio-temporal-data.html) | [**Exercise 5**](exercises/exercise_05.pdf) | 
| 11 | 2020-01-12 | [**Combining rasters and vector layers**](combining-rasters-and-vector-layers.html) | | 
| 12 | 2020-01-19 | [**Spatial interpolation of point data**](spatial-interpolation-of-point-data.html) | [**Exercise 6**](exercises/exercise_06.pdf) | 

## Exercise submission dates

| Exercise | Date |
|:--:|:---:|
| 01 | 2019-11-17 |
| 02 | 2019-12-01 |
| 03 | 2019-12-15 |
| 04 | 2020-01-05 |
| 05 | 2020-01-19 |
| 06 | 2020-02-02 |

## Exercise instructions

### Guidelines

* Submission in through **Moodle** only
* The submission should be a **single R code file**
* File name is your **last name** + **exercise number**, such as `dorman_01.R`
* The code needs to run **as is**, assuming that - 
    * All required packages are **installed**
    * All data files are in the user's **working directory**
* **Do not** include `install.packages` expressions
* **Do** include `library` expressions
* **Do not** include `setwd` expressions
* Self-check - 
    * Start a new R **session**
    * **Manually** set the **working directory** to where the data files are
    * **Run** the entire script
* **Late** submission up to -20% and additional -10% per week, unless coordinated in advance

```{r, eval=FALSE}
# Exercise 01
# Michael Dorman

library(raster)

# Question 01
r = raster("elevation.tif")
plot(r)

# Question 02
...
```


### Generality

* The code needs to be as **general** as possible - 
    * Do not use **specific values** except the ones given in the question
    * When there are **intermediate results** - assign them to a variable and use the variable, not the value

* Question: what is the last value of `c(5,3,1,7,2,3,2,6)`?
* Good solution - 

```{r}
x = c(5,3,1,7,2,3,2,6)
x[length(x)]
```

* Bad solution - 

```{r}
x = c(5,3,1,7,2,3,2,6)
x[8]
```

### Brackets

* Use **brackets** to make sure the right order of operations is done

* Question: print the subset of the last four values in `c(5,3,1,7,2,3,2,6)`
* Correct solution - 

```{r}
x = c(5,3,1,7,2,3,2,6)
x[(length(x)-3):length(x)]
```

* Wrong solution - 

```{r}
x = c(5,3,1,7,2,3,2,6)
x[length(x)-3:length(x)]
```

### Required value or output

* Note which **outputs** or **printouts** are requested

### Packages

* The `sp`, `rgeos`, `rgdal` and `raster` cannot be used in the exercises. 

## Other resources

### Tutorials

* Adrian Baddeley, 2008. **Analysing spatial point patterns in R**. [[**PDF**](resources/PointPatterTutorial.pdf)]

### Papers

* Pebesma, E., Bivand, R. S. (2005). **Classes and Methods for Spatial Data: the sp Package**. R news, 5(2), 9-13. [[**PDF**](resources/Rnews_2005-2.pdf)]
* Pebesma, E. (2018). **Simple Features for R: Standardized Support for Spatial Vector Data**. The R Journal, 10(1):439-446. [[**PDF**](resources/RJ-2018-009.pdf)]

### Books - General

* Murrell, P. (2010). **Introduction to Data Technologies**. Chapman and Hall/CRC. [[**PDF**](resources/Murrell 2010 Introduction to Data Technologies.pdf)] [[**Website**](https://www.stat.auckland.ac.nz/~paul/ItDT/)]
* Wickham, H. (2014). **Advanced R**. Chapman and Hall/CRC. [[**HTML**](http://adv-r.had.co.nz/)]

### Books - Spatial data

* Hengl, T. (2009). **A Practical Guide to Geostatistical Mapping**. [[**PDF**](resources/Hengl 2009 A Practical Guide to Geostatistical Mapping.pdf)]
* Lovelace, R., Nowosad, J., Muenchow, J. (2019). **Geocomputation with R**. Chapman and Hall/CRC. [[**HTML**](https://geocompr.robinlovelace.net/)]
* Pebesma, E., Bivand, R. (in preparation). **Spatial Data Science**. [[**HTML**](https://keen-swartz-3146c4.netlify.com/)]
